---
title: <span style="color:green;font-size:34pt">The Value of Transmission from Pacific Northwest to California</span>
subtitle: <span style="font-size:20pt">An Analysis Based on CAISO Day Ahead and 15 Minute Market Prices</span>
author: Elena Stolpovsky
date: August 25, 2021
output: 
  ioslides_presentation: 
    css: styles.css
    smaller: true
---


```{r setup, include=F}
knitr::opts_chunk$set(echo = T,message=FALSE,warning=FALSE,widgetframe_widgetsdir = 'widgets')
library(visNetwork, quietly = TRUE)
library(widgetframe)
```

## Physical Transmission Rights 

A physical transmission right (PTR) gives the holder the exclusive right to use a particular interconnection *in one direction* to transfer a *predefined quantity of energy* *from one market hub to the other*.

- Market Hubs: Mid-Columbia (MIDC) and North of Path 15 (NP15)
- Direction: North to South
- Quantity: 1 MW (can be rescaled for a different transmission capacity)

Who needs to understand the value of PTR:

- Utilities in the PNW or generation owners that use the PNW-California path
- Owners of PTRs (e.g. BPA) that need to price PTR contracts
- Regulators considering building new transmission

## Value from Owning Transmission

Opportunity for arbitrage between the market hubs.

Arbitrage can occur in different markets:

- Long-term forward market (typically peak and off-peak prices)
- Day-ahead market (hourly prices)
- Real time market

  - 15-minute market
  - 5-minute market


Data:

- Source: CAISO OASIS day ahead and 15-min market prices at MIDC (PGE share) and NP15
- Period: August 1 2019 midnight - July 31 2021 11am.
- Frequency: hourly for day ahead data, by 15 minutes for 15-min market data


<!--  
### Value of Physical Transmission Rights  

A PTR gives the holder the **option** to transfer energy from one market hub to another.

A purchase of a PTR is analogous to a purchase of an option:
  - Upfront cost
  - opportunity for nonnegative payoffs in the future


## Loading the CAISO OASIS Data

CAISO OASIS website allows to download at most 1 month of price data


```{r, eval=FALSE}
file_list <- paste(path, list.files(path), sep = "") 
#list of files with monthly prices

fmm <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files
  fmm <-
    rbindlist(list(fmm, temp), use.names = T) #for each iteration, 
  #bind the new data to the building dataset
}
```
-->

<!--

## CAISO Day-Ahead Market

- The utility is able to commit resources and execute bilateral trades in the day ahead. 
  - More certainty that it will be able to advantage of the price spread
- This provides a lower bound  
-->

## Day-Ahead Prices at MIDC and NP15, August 2019

<iframe src="damprices.html"></iframe>

<!--
## Summary of Day-Ahead Prices 

<iframe src="damsummary.html"></iframe>
-->

## Day Ahead Market Price Sread: Price at NP15 - Price at MIDC

<iframe src="damspread.html"></iframe>

## 15-Min Market Pices at MIDC and NP15, August 2019

<iframe src="fmmprices.html"></iframe>

## 15-Min Market Price Sread: Price at NP15 - Price at MIDC

<iframe src="fmmspread.html"></iframe>

## DAM and FMM Prices and Spreads Between NP15 and MIDC, August 2019 - August 2021

<iframe src="allmarketsummary.html"></iframe>

## Nonnegative DAM and FMM Price Spreads

Nonnegative price spreads between NP15 and MIDC represent the value of the option to transmit energy south between these hubs.

The value of 1 MW of PTR from MIDC to NP15 in a given hour =  (nonnegative price spread per hour in $/MWh)*1 MW, adjusted for transmission losses.

<!--Transforming 15-Minute Nonnegative Spreads into Hourly Values-->
I average 15-minute nonnegative spreads by hour to obtain **hourly nonnegative price spreads** in the FMM.

<!--
```{r, eval=F}
#Averaging 15-min nonnegative price spreads into hourly nonnegative price spreads
fmmhourly <-
  fmmpivot %>% mutate(hourlystart = strptime(start, format = "%Y-%m-%d %H")) %>%
  group_by(hourlystart) %>% summarize(nonnegative_spread =
                            mean(nonnegative_spread)) 
```
-->
<iframe src="fmmhourlysummary.html"></iframe>

## Nonnegative Price Spreads in DAM and FMM in August 2019

<iframe src="dam_rtm_nonnegative_spreads.html"></iframe>


## Seasonality of Nonnegative Spreads in the DAM and FMM

- High hourly Seasonality in the DAM and FMM
  - Morning and evening peaks are sharper in California than in the Pacific Northwest.
- No weekly seasonality
  - The day of the week does not appear to affect price spreads
- Possible monthly seasonality
  - 15-min market spreads were highest in July

## Nonnegative Price Spreads by Hour of the Day

<iframe src="hourly_shape_of_spreads.html"></iframe>

## Nonnegative Price Spreads by Day of the Week

<iframe src="weekday_shape_of_spreads.html"></iframe>

## Nonnegative Price Spreads by Month of the Year

<iframe src="monthly_shape of _spreads.html"></iframe>

## Valuation Based on the NPV of Payoffs

Model assumptions and inputs:

- Transmission can always be utilized at full capacity
- Transmission losses = 5% ([eia.gov](https://www.eia.gov/tools/faqs/faq.php?id=105&t=3]))
- Yearly discount rate = 10% (based on the company's cost of capital)

$NPV= \sum_{t=0}^{n}\frac{R_t}{(1+i)^t}$

$t$: hours

hourly payoffs: $R_t =$ nonnegative hourly price spread * $0.95$ 

monthly discounting: $i=10\%/12$

The NPV (as of August 1 2019) of utilizing 1 MW of transmission from PNW to CA for 2 years in the **DAM** is **$23,000**.

The NPV of utilizing 1 MW of transmission for 2 years in the **FMM** is **$177,000**.

## 10 Year Valuation

PTR contracts may be long-term.

I create a "naive" forecast of nonnegative price spreads in the day ahead and real time markets for 2021 - 2029:
  price spreads for every 2 year forecast period until July 2029 = price spreads in 2019 - 2021
  
```{r, eval=F}
naive_forecast_until_aug1_2029 <-rep(fmmhourly$nonnegative_spread, times = 5)
```


The NPV (as of August 1 2019) of utilizing 1 MW of transmission from PNW to CA for 10 years in the **DAM** is **$80,000**.

The NPV of utilizing 1 MW of transmission for 10 years in the **FMM** is **$609,000**.

The 15-min market brings **7.6 times** more arbitrage value from the difference in MIDC and Northern CA prices.


## Quantifying Uncertainty 

Can model 2 price series separately or model the spread directly.

  - If modeling 2 price series, need to preserve the correlation between the two price points 
  
Price spreads exhibit characteristics similar to other energy price series:

- Mean reverting (models like exponential smoothing)
- Frequent spikes or jumps
- High volatility, or fat tails

  
To quantify uncertainty

- use the model to generate N (e.g. 1000) simulations of the series for the valuation period
- calculate N NPVs
- Obtain summary statistics from the distribution of NPVs: mean, standard errors, and percentiles.


 

## Quantifying Uncertainty. A Bootstrap Approach.

Simulate nonnegative price spreads in a given hour of the day by sampling from historical data for this hour. 

If N is the number of simulations needed,
Y is the number of years of data to be simulated:

- Generate a sample of hour 0 data:
  - Take a sample with replacement of size $N*Y*$(number of hours in a year) from hour 0 historical data. 

```{r, eval=F}
sample(hour0, size = N*Y*8765, replace = TRUE)
```
- Samples of hour 1 to hour 23 data are generated in the same way.
- Transform the samples into N simulations of hourly data for Y years.
- For each simulation, calculate the NPV.
- Obtain summary statistics from the distribution of NPVs.





## Discussion

Other Sources of Value from PTRs:

- Sale of Renewable Energy Credits
- Reduction of curtailment of wind and solar energy.
  - Increase the value of renewable energy projects and promote their development

Considerations:

- Can the utilities always take full advantage of price spreads in the 15-minute market?
- How does the participation in the 5-minute market affect the value of transmission?
  - I expect greater option value, but lower market volume and greater volatility.
  
# Questions?

## Loading the CAISO OASIS Data

CAISO OASIS website allows to download at most 1 month of price data


```{r, eval=FALSE}
file_list <- paste(path, list.files(path), sep = "") 
#list of files with monthly prices

fmm <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files
  fmm <-
    rbindlist(list(fmm, temp), use.names = T) #for each iteration, 
  #bind the new data to the building dataset
}
```
