---
title: <span style="color:green;font-size:34pt">The Value of Transmission from Pacific Northwest to California</span>
subtitle: <span style="font-size:20pt">An Analysis Based on CAISO Day Ahead and 15 Minute Market Prices</span>
author: Elena Stolpovsky, PhD
date: August 25, 2021
output: 
  ioslides_presentation: 
    css: styles.css
    smaller: true
---


```{r setup, include=F}
knitr::opts_chunk$set(
  echo = T,
  message = FALSE,
  warning = FALSE,
  widgetframe_widgetsdir = 'widgets'
)
library(visNetwork, quietly = TRUE)
library(widgetframe)
options(scipen = 999,  digits = 2)
Sys.setenv(TZ = 'GMT')
```

## Physical Transmission Rights 

A physical transmission right (PTR) gives the holder the exclusive right to use a particular interconnection *in one direction* to transfer a *predefined quantity of energy* *from one market hub to the other*.

- Market Hubs: Mid-Columbia (MIDC) and North of Path 15 (NP15)
- Direction: North to South
- Quantity: 1 MW (can rescaled for a different transmission capacity)

Who needs to understand the value of PTR:

- Utilities in the PNW or generation owners that use the PNW-California path
- Owners of PTRs (e.g. BPA) that need to price PTR contracts
- Regulators considering building new transmission

## Value from Owning Transmission

Opportunity for arbitrage between the market hubs.

Arbitrage can occur in different markets:

- Long-term forward market (typically peak and off-peak prices)
- Day-ahead market (hourly prices)
- Real time market

  - 15-minute market
  - 5-minute market


Data:

- Source: CAISO OASIS day ahead and 15-min market prices at MIDC (PGE share) and NP15
- Period: August 1 2019 midnight - July 31 2021 11am.
- Frequency: hourly for day ahead data, by 15 minutes for 15-min market data

<!--

## Loading the CAISO OASIS Data

CAISO OASIS website allows to download at most 1 month of price data


```{r, eval=FALSE}
file_list <- paste(path, list.files(path), sep = "") 
#list of files with monthly prices

fmm <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files
  fmm <-
    rbindlist(list(fmm, temp), use.names = T) #for each iteration, 
  #bind the new data to the building dataset
}
```
 
-->

## Day-Ahead Prices at MIDC and NP15

```{r, echo=F}
library(data.table)
library(tidyverse)
library(htmlwidgets)
library(DT)
path = "C:/elena/caiso_prices/analysis/data/DAM/"
file_list <- paste(path, list.files(path), sep = "")

dam <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files using the fread function from the data.table package
  dam <-
    rbindlist(list(dam, temp), use.names = T) #for each iteration, bind the new data to the building dataset
}
dam <- dam %>% distinct()

names(dam) <-
  c(
    'start',
    'end',
    'opdt',
    'ophr',
    'opint',
    'nodidxml',
    'nodeid',
    'node',
    'marketrun',
    'lmp_type',
    'xmldataitem',
    'pnoderesmrid',
    'grptype',
    'pos',
    'price',
    'group'
  )
dam <-
  dam %>% filter(lmp_type == "LMP",
                 node %in% c('TH_NP15_GEN-APND', 'MIDC_5_PGESHARE-APND')) %>% select(start, end, node, price) %>% mutate(node =
                                                                                                                           case_when(
                                                                                                                             node == 'MIDC_5_PGESHARE-APND' ~ 'midc',
                                                                                                                             node == 'TH_NP15_GEN-APND' ~ 'np15'
                                                                                                                           )) %>% arrange(start) %>% mutate(start = start - 7 * 60 * 60, end = end -
                                                                                                                                                              7 * 60 * 60) %>% # correct for the fact that Pacific time was read as GMT
  filter(start < "2021-08-01 UTC")
#choose 2 years of data exactly

library(plotly)
dam10days <- dam %>% filter(start < "2019-08-10 07:00:00 UTC")

p <-
  ggplot(dam10days, aes(start, price, color = node)) + geom_line() + xlab("Starting Interval Time") +
  ylab("Price, $/MWh") + labs(color = '') + labs(color = '')

ggplotly(p, height = 400, width = 700)

```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019</span>


## Day Ahead Market Price Sread: Price at NP15 - Price at MIDC

```{r, echo=F}
dampivot <-
  dam %>% pivot_wider(names_from = node, values_from = price) %>% mutate(spread =
                                                                           np15 - midc) %>% mutate(nonnegative_spread = (spread > 0) * spread)

min <- dampivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ min(.))
mean <-
  dampivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ mean(.))

max <- dampivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ max(.))
sd <- dampivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ sd(.))
damsum <-
  rbind(min, mean, max, sd) %>% round(1) %>% t() %>% as.data.frame()
names(damsum) = c('Min', 'Mean', 'Max', 'St.Dev')
rownames(damsum) <-
  c('DAM MIDC Price',
    'DAM NP15 Price',
    'DAM Spread',
    'DAM Nonnegative Spread')

damtab <- datatable(damsum, options = list(dom = 't'), width = .7)

dampivot10days <-
  dampivot %>% filter(start < "2019-08-10 07:00:00 UTC")
p <-
  ggplot(dampivot10days, aes(start)) + geom_line(aes(y = spread)) + geom_hline(yintercept =
                                                                                 0, color = "red") + xlab("Starting Interval Time") + ylab("Price Spread, $/MWh")
ggplotly(p, height = 400, width = 700)

```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019</span>

## 15-Min Market Pices at MIDC and NP15

```{r, echo=F}
path = "C:/elena/caiso_prices/analysis/data/FMM/"
file_list <- paste(path, list.files(path), sep = "")

fmm <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files using the fread function from the data.table package
  fmm <-
    rbindlist(list(fmm, temp), use.names = T) #for each iteration, bind the new data to the building dataset
}
fmm <- fmm %>% distinct()
names(fmm) <-
  c(
    'start',
    'end',
    'opdt',
    'ophr',
    'nodidxml',
    'nodeid',
    'node',
    'marketrun',
    'lmp_type',
    'xmldataitem',
    'pnoderesmrid',
    'grptype',
    'pos',
    'price',
    'opinterval',
    'group'
  )

fmm <-
  fmm %>% filter(lmp_type == "LMP",
                 node %in% c('TH_NP15_GEN-APND', 'MIDC_5_PGESHARE-APND')) %>% select(start, end, node, price) %>% mutate(node =
                                                                                                                           case_when(
                                                                                                                             node == 'MIDC_5_PGESHARE-APND' ~ 'midc',
                                                                                                                             node == 'TH_NP15_GEN-APND' ~ 'np15'
                                                                                                                           )) %>% filter(node %in% c('midc', 'np15')) %>% mutate(start = start - 7 *
                                                                                                                                                                                   60 * 60, end = end - 7 * 60 * 60) %>% # correct for the fact that Pacific time was read as GMT <-
  filter(start < "2021-08-01 UTC") %>% arrange(start)

fmm10days <- fmm %>% filter(start < "2019-08-10 07:00:00 UTC")

p <-
  ggplot(fmm10days, aes(start, price, color = node)) + geom_line() + xlab("Starting Interval Time") +
  ylab("Price, $/MWh") + labs(color = '')
ggplotly(p, height = 400, width = 700)

```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 </span>


## 15-Min Market Price Sread: Price at NP15 - Price at MIDC

```{r, echo=F, results = FALSE}
fmmpivot <-
  fmm %>% pivot_wider(names_from = node, values_from = price) %>% mutate(spread =
                                                                           np15 - midc) %>% mutate(nonnegative_spread = (spread > 0) * spread)

min <- fmmpivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ min(.))
mean <-
  fmmpivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ mean(.))

max <- fmmpivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ max(.))
sd <- fmmpivot %>% select(midc, np15, spread, nonnegative_spread) %>%
  summarise_all( ~ sd(.))
fmmsum <-
  rbind(min, mean, max, sd) %>% round(1) %>% t() %>% as.data.frame()
names(fmmsum) = c('Min', 'Mean', 'Max', 'St.Dev')
rownames(fmmsum) <-
  c(
    '15-min Market MIDC Price',
    '15-min Market NP15 Price',
    '15-min Market Spread',
    '15-min Market Nonnegative Spread'
  )

fmmtab <- datatable(fmmsum, options = list(dom = 't'), width = .7)

allmarketsum <- rbind(damsum, fmmsum)
setDT(allmarketsum, keep.rownames = TRUE)[]
names(allmarketsum)[1] <- 'Series'

```


```{r, echo=F}

fmmpivot10days <-
  fmmpivot %>% filter(start < "2019-08-10 07:00:00 UTC")
p <-
  ggplot(fmmpivot10days, aes(start)) + geom_line(aes(y = spread)) + geom_hline(yintercept =
                                                                                 0, color = "red") + xlab("Starting Interval Time") + ylab("Price Spread, $/MWh")

ggplotly(p, height = 400, width = 700)

```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 </span>

## DAM and FMM Prices and Spreads Between NP15 and MIDC, 2019 - 2021

```{r, echo=F}
allmarkettab <-
  datatable(allmarketsum, options = list(dom = 't'), width = .5) %>% formatStyle('Series', "white-space" =                                                                                   "nowrap")
allmarkettab
```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 - 07/2021</span>



## The Series of Interest: Nonnegative DAM and FMM Price Spreads

Nonnegative price spreads between NP15 and MIDC represent the value of the option to transmit energy south between these hubs.

The value of 1 MW of PTR from MIDC to NP15 in a given hour =  (nonnegative price spread per hour in $/MWh)*1 MW, adjusted for transmission losses.

I average 15-minute nonnegative spreads by hour to obtain **hourly nonnegative price spreads** in the FMM.

```{r, echo=F}
fmmhourly <-
  fmmpivot %>% mutate(hourlystart = strptime(start, format = "%Y-%m-%d %H")) %>% group_by(hourlystart) %>% summarize(nonnegative_spread =
                                                                                                                       mean(nonnegative_spread)) %>% ungroup() %>% filter(!is.na(hourlystart)) %>% mutate(hourlystart = as.POSIXct(hourlystart))

min <- fmmhourly %>% select(nonnegative_spread) %>%
  summarise_all( ~ min(.))
mean <- fmmhourly %>% select(nonnegative_spread) %>%
  summarise_all( ~ mean(.))
max <- fmmhourly %>% select(nonnegative_spread) %>%
  summarise_all( ~ max(.))
sd <- fmmhourly %>% select(nonnegative_spread) %>%
  summarise_all( ~ sd(.))
sum <-
  rbind(min, mean, max, sd) %>% round(1) %>% t() %>% as.data.frame()
names(sum) = c('Min', 'Mean', 'Max', 'St.Dev')
rownames(sum) <- 'FMM Hourly Nonnegative Spread'

fmmhourlytab <- datatable(sum, options = list(dom = 't'), width = .2)

fmmhourlytab
```

<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 - 07/2021</span>

## Nonnegative Hourly Price Spreads in DAM and FMM in August 2019

```{r, echo=F}
damrtm <-
  dampivot %>% left_join(fmmhourly, by = c("start" = "hourlystart")) %>% select(start, spread_dam =
                                                                                  nonnegative_spread.x, spread_fmm = nonnegative_spread.y) %>% filter(!is.na(spread_fmm))

damrtm$average_fmm_spread <- mean(damrtm$spread_fmm)
p <-
  ggplot(damrtm[1:240, ], aes(start)) + geom_line(aes(y = spread_dam, color =
                                                        "spread dam")) +
  geom_line(aes(y = spread_fmm, color = "spread fmm")) + xlab("Starting Interval Time") +
  ylab("Price Spread, $/MWh") + labs(color = '')

#+geom_line(aes(y=average_fmm_spread,color="average_fmm_spread"))
ggplotly(p, height = 400, width = 700)

```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019</span>

## Seasonality of Nonnegative Spreads in the DAM and FMM

- High hourly Seasonality in the DAM and FMM
  - Morning and evening peaks are sharper in California than in the Pacific Northwest.
- No weekly seasonality
  - The day of the week does not appear to affect price spreads
- Possible monthly seasonality
  - 15-min market spreads were highest in July

## Nonnegative Price Spreads by Hour of the Day


```{r, echo=F}
damrtm_by_day <-
  damrtm %>% group_by(hour = hour(start)) %>% summarise(
    spread_d = mean(spread_dam),
    sd_dam = sd(spread_dam),
    spread_f = mean(spread_fmm),
    sd_fmm = sd(spread_fmm)
  ) %>% mutate(spread_dam = spread_d, spread_fmm = spread_f)

p <-
  ggplot(damrtm_by_day, aes(hour)) + geom_line(aes(y = spread_dam, color =
                                                     "spread dam")) +
  geom_line(aes(y = spread_fmm, color = "spread fmm")) + geom_line(aes(y =
                                                                         sd_dam, color = "sd dam"),
                                                                   linetype = "dashed",
                                                                   alpha = 0.7) +
  geom_line(aes(y = sd_fmm, color = "sd fmm"),
            linetype = "dashed",
            alpha = 0.7) + xlab("Starting Interval Time") + ylab("Price Spread, $/MWh") +
  labs(color = '')
ggplotly(p, height = 400, width = 700)
```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 - 07/2021</span>

## Nonnegative Price Spreads by Day of the Week


```{r, echo=F}
damrtm_by_week <-
  damrtm %>% group_by(weekday = wday(start)) %>% summarise(
    spread_d = mean(spread_dam),
    sd_dam = sd(spread_dam),
    spread_f = mean(spread_fmm),
    sd_fmm = sd(spread_fmm)
  ) %>% mutate(spread_dam = spread_d, spread_fmm = spread_f)

p <-
  ggplot(damrtm_by_week, aes(weekday)) + geom_line(aes(y = spread_dam, color =
                                                         "spread dam")) +
  geom_line(aes(y = spread_fmm, color = "spread fmm")) + geom_line(aes(y =
                                                                         sd_dam, color = "sd dam"),
                                                                   linetype = "dashed",
                                                                   alpha = 0.7) +
  geom_line(aes(y = sd_fmm, color = "sd fmm"),
            linetype = "dashed",
            alpha = 0.7) + xlab("Day of the Week") + ylab("Price Spread, $/MWh") +  labs(color =
                                                                                           '')
ggplotly(p, height = 400, width = 700)
```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 - 07/2021</span>

## Nonnegative Price Spreads by Month of the Year


```{r, echo=F}
damrtm_by_month <-
  damrtm %>% group_by(month = month(start)) %>% summarise(
    spread_d = mean(spread_dam),
    sd_dam = sd(spread_dam),
    spread_f = mean(spread_fmm),
    sd_fmm = sd(spread_fmm)
  ) %>% mutate(spread_dam = spread_d, spread_fmm = spread_f)

p <-
  ggplot(damrtm_by_month, aes(month)) + geom_line(aes(y = spread_dam, color =
                                                        "spread dam")) +
  geom_line(aes(y = spread_fmm, color = "spread fmm")) + geom_line(aes(y =
                                                                         sd_dam, color = "sd dam"),
                                                                   linetype = "dashed",
                                                                   alpha = 0.7) +
  geom_line(aes(y = sd_fmm, color = "sd fmm"),
            linetype = "dashed",
            alpha = 0.7) + xlab("Month of the Year") + ylab("Price Spread, $/MWh") +
  labs(color = '')
ggplotly(p, height = 400, width = 700)
```
<span style="font-size:10pt">*Source:* [CAISO OASIS](http://oasis.caiso.com/mrioasis/logon.do), 08/2019 - 07/2021</span>

## Valuation Based on the NPV of Payoffs

Model assumptions and inputs:

- Transmission can always be utilized at full capacity
- Transmission losses = 5% ([eia.gov](https://www.eia.gov/tools/faqs/faq.php?id=105&t=3]))
- Yearly discount rate = 10% (based on the company's cost of capital)

$NPV= \sum_{t=0}^{n}\frac{R_t}{(1+i)^t}$

$t$: hours

hourly payoffs: $R_t =$ nonnegative hourly price spread * $0.95$ 

monthly discounting: $i=10\%/12$

The NPV (as of August 1 2019) of utilizing 1 MW of transmission from PNW to CA for 2 years in the **DAM** is **$23,000**.

The NPV of utilizing 1 MW of transmission for 2 years in the **FMM** is **$177,000**.

## 10 Year Valuation

PTR contracts may be long-term.

I create a "naive" forecast of nonnegative price spreads in the day ahead and real time markets for 2021 - 2029:
  price spreads for every 2 year forecast period until July 2029 = price spreads in 2019 - 2021.
  
```{r, eval=F}
naive_forecast_until_aug1_2029 <-rep(fmmhourly$nonnegative_spread, times = 5)
```


The NPV (as of August 1 2019) of utilizing 1 MW of transmission from PNW to CA for 10 years in the **DAM** is **$80,000**.

The NPV of utilizing 1 MW of transmission for 10 years in the **FMM** is **$609,000**.

The 15-min market brings **7.6 times** more arbitrage value from the difference in MIDC and Northern CA prices.


## Quantifying Uncertainty 

Can model 2 price series separately or model the spread directly.

  - If modeling 2 price series, need to preserve the correlation between the two price points 
  
Price spreads exhibit characteristics similar to other energy price series:

- Mean reverting (models like exponential smoothing)
- Frequent spikes or jumps
- High volatility, or fat tails

  
To quantify uncertainty

- use the model to generate N (e.g. 1000) simulations of the series for the valuation period
- calculate N NPVs
- Obtain summary statistics from the distribution of NPVs: mean, standard errors, and percentiles.


 

## Quantifying Uncertainty. A Bootstrap Approach.

Simulate nonnegative price spreads in a given hour of the day by sampling from historical data for this hour. 

If N is the number of simulations needed,
Y is the number of years of data to be simulated:

- Generate a sample of hour 0 data:
  - Take a sample with replacement of size $N*Y*$(number of hours in a year) from hour 0 historical data. 

```{r, eval=F}
sample(hour0, size = N*Y*8765, replace = TRUE)
```
- Samples of hour 1 to hour 23 data are generated in the same way.
- Transform the samples into N simulations of hourly data for Y years.
- For each simulation, calculate the NPV.
- Obtain summary statistics from the distribution of NPVs.





## Discussion

Other Sources of Value from PTRs:

- Sale of Renewable Energy Credits
- Reduction of curtailment of wind and solar energy.
  - Increase the value of renewable energy projects and promote their development

Considerations:

- Can the utilities always take full advantage of price spreads in the 15-minute market?
- How does the participation in the 5-minute market affect the value of transmission?
  - I expect greater option value, but lower market volume and greater volatility.
  
# Questions?


## Extra Slide: Loading the CAISO OASIS Data

CAISO OASIS website allows to download at most 1 month of price data. I use the following code to read in all monthly files and concatenate them into a single file. 


```{r, eval=FALSE}
file_list <- paste(path, list.files(path), sep = "") 
#list of files with monthly prices

fmm <- data.frame()
for (i in 1:length(file_list)) {
  temp <-
    fread(file_list[i]) #read in files
  fmm <-
    rbindlist(list(fmm, temp), use.names = T) #for each iteration, 
  #bind the new data to the building dataset
}
```
-->